Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-12-28 paper.tex
Added some more of Barry's edits.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\abs{\hat{a} - a_0}^2$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\hat{\theta} - \theta_0\rangle_\pi^2$ and $(\hat{\rho} - \rho_0)^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\hat{\theta} - \theta_0\rangle_\pi^2$ rather than $(\hat{\theta} - \theta_0)^2$ is the appropriate measure of error for the phase parameter.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi^2$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho$ is a nonnegative real.  Now,
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $\abs{P}$ and $\abs{D}$ increase in such a way that
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$,
\begin{enumerate}
\item $\sfracpart{\hat{\theta} - \theta_0}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:SSallparams} we obtain $SS$ conditioned on minimisation with respect to the data symbols,
 \begin{align*}
SS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$ conditioned on minimisation with respect to the data symbols, 
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)),
\end{equation}
where $\Re(\cdot)$ denotes the real part.  Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  Substituting this expression into $SS(\rho, \theta)$ gives $SS$ conditioned upon minimisation with respect to $\rho$ and the data symbols,
\[
SS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]

We are interested in analysing $\hat{\theta}$, the minimiser of $SS(\theta)$, or equivalently, the maximiser of $\Re(Z(\theta))$.  Recalling the definition of $R_i$ and $\Phi_i$ from~\eqref{eq:RiandPhii},
\begin{align*}
y_i &= a_0 s_i + w_i \\
&= a_0 s_i \left( 1 + \frac{w_i}{a_0 s_i} \right) \\
&= a_0 s_i R_i e^{j \Phi_i} \\
&= \rho_0 R_i e^{j ( \Phi_i + \theta_0 + \angle{s_i}) }.
\end{align*}
%The $\Phi_i$ can be identified as \emph{circular~random~variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  
Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta},
\begin{align*}
\hat{u}_i(\theta) &= \round{\angle{y_i} - \theta} \\
&= \round{\theta_0 + \Phi_i + \angle{s_i} - \theta} \\
&\equiv \round{ \fracpart{\theta_0 - \theta}_{\pi} + \Phi_i + \angle{s_i}} \pmod{2\pi} \\
&= \round{ \lambda + \Phi_i + \angle{s_i} },
\end{align*}
where we put $\lambda = \fracpart{\theta_0 - \theta}_\pi$.  Because $\hat{d}_i^*(\theta) = e^{-j\hat{u}_i(\theta)}$, it follows that, when $i \in D$,
\begin{align}
 y_i e^{-j\theta} \hat{d}_i^*(\theta) &= \rho_0 R_i e^{j(\lambda + \Phi_i + \angle{s_i} - \round{\lambda + \Phi_i + \angle{s_i}})} \nonumber \\
&= \rho_0 R_i e^{j(\lambda + \Phi_i - \round{\lambda + \Phi_i})} \nonumber  \\
&= \rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}} \label{eq:yethetadhat}
\end{align}
since $\round{x + \angle{s_i}} = \round{x} + \angle{s_i}$ for all $x \in \reals$ as a result of $\angle{s_i}$ being a multiple of $\tfrac{2\pi}{M}$.  Otherwise, when $i \in P$,  
\[
y_i e^{-j\theta} p_i^* = \rho_0 R_i e^{j(\lambda + \Phi_i)}.
\]
Now,
\[
Z(\theta) = \rho_0 \sum_{i \in P} R_i e^{j(\lambda + \Phi_i)} + \rho_0  \sum_{i \in D} R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
and
\[
\Re(Z(\theta)) = \rho_0 \sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \rho_0 \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}.
\] 
Let 
\begin{equation}\label{eq:GLdefn}
G_L(\lambda) = \frac{1}{\rho_0 L} \Re(Z(\theta))
\end{equation}
and put $\hat{\lambda}_L = -\sfracpart{\hat{\theta} - \theta_0}_\pi = \sfracpart{\theta_0 - \hat{\theta}}_\pi$.  Since $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ then  $\hat{\lambda}_L$ is the maximiser of $G_L(\lambda)$.  We will show that $\hat{\lambda}_L$ converges almost surely to zero as $L \rightarrow \infty$.  The proof of part 1 of Theorem~\ref{thm:consistency} follows from this.

Recall the functions $G$, $h_1$ and $h_2$ defined in the statement of Theorem~\ref{thm:consistency}.  Observe that
\[
\expect G_L(\lambda) = \frac{\abs{P}}{L} h_1(\lambda) + \frac{\abs{D}}{L} h_2(\lambda)
\]
and since $\frac{\abs{P}}{L} \rightarrow p$ and $\frac{\abs{D}}{L} \rightarrow d$ as $L \rightarrow \infty$, then
\[
\lim_{L \rightarrow\infty} \expect G_L(\lambda) = G(\lambda) = p h_1(\lambda)   +  d h_2(\lambda).
\]
As is customary, let $(\Omega, \mathcal F, \prob)$ be the probability space over which the random variables $\{w_i\}$ are defined.  Let $A$ be the subset of the sample space $\Omega$ upon which $G(\hat{\lambda}_L) \rightarrow G(0)$ as $L\rightarrow\infty$.  Lemma~\ref{lem:convtoexpGlamL} shows that $\prob\{A\} = 1$.  Let $A'$ be the subset of the sample space upon which $\hat{\lambda}_L \rightarrow 0$ as $L\rightarrow \infty$.  Because $G(x)$ is uniquely maximised at $x=0$, it follows that $G(\hat{\lambda}_L) \rightarrow G(0)$ only if $\hat{\lambda}_L \rightarrow 0$ as $L \rightarrow\infty$. So $A \subseteq A'$ and therefore $\prob\{A'\} \geq \prob\{A\} = 1$.  Part 1 of Theorem~\ref{thm:consistency} follows.

@@

\begin{lemma}\label{lem:empiricprocforrho} Let $T_L(\lambda) = \expect G_L(\lambda)$.  We have
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) \big) = o_P(1) + X_L,
\]
where $X_L = \sqrt{L} \big( G_L(0) - T_L(0) \big)$.
\end{lemma}
\begin{IEEEproof}
Write
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) \big) = Y_L(\hat{\lambda}_L) + X_L
\]
where
\begin{equation}\label{eq:YLdef}
Y_L(\lambda) = \sqrt{L}\big( G_L(\lambda) - T_L(\lambda) \big) - X_L.
\end{equation}
is an empirical process indexed by $\lambda$, similar to $W_L$ from~\eqref{eq:WLdef}.  As with $W_L$ results from the literature on empirical processes can be used to show that  
%Lemma~\ref{lem:ZYLempiricproc} 
%in Section~\ref{sec:an-empirical-process} 
%shows that 
for any $\delta > 0$ and $\nu > 0$, there exists $\epsilon > 0$ such that
\[
\Pr\left\{ \sup_{\sabs{\lambda}<\epsilon}\sabs{Y_L(\lambda)} > \delta  \right\} < \nu.
\]
The proof now follows by an argument analogous to that in Lemma~\ref{lem:empiricprocc}.
\end{IEEEproof}

>>>>

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\sabs{\hat{a} - a_0}$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\hat{\theta} - \theta_0\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\hat{\theta} - \theta_0\rangle_\pi$ rather than $\hat{\theta} - \theta_0$ is the appropriate measure of error for the phase parameter.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain $SS$ minimised with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:SSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho$ is a nonnegative real.  Now,
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
We have slightly abused notation here by reusing $SS$. This should not cause confusion as $SS(a, \{d_i, i \in D\})$, $SS(\rho, \theta, \{d_i, i \in D\})$, and $SS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $\abs{P}$ and $\abs{D}$ increase in such a way that
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$ then
\begin{enumerate}
\item $\sfracpart{\hat{\theta} - \theta_0}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:SSallparams} we obtain $SS$ minimised with respect to the data symbols,
 \begin{align*}
SS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$,
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)),
\end{equation}
where $\Re(\cdot)$ denotes the real part.  %Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  
Substituting this expression into $SS(\rho, \theta)$ gives $SS$ minimised with respect to $\rho$ and the data symbols,
\[
SS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
We again abuse notation by reusing $SS$, but this should not cause confusion as $SS(\rho,\theta)$ and $SS(\theta)$ are easily told apart by their inputs.  By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  %In this case we would replace $\theta$ with $\fracpart{\theta + \phi}$ and $\hat{\rho}(\theta)$ with $-\hat{\rho}(\theta)$ and observe that
%\[
%SS(\rho,\theta) = SS(-\rho, \fracpart{\theta + \pi})
%\]
The least square estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $SS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently
\[
\hat{\theta} = \arg\max_{\theta \in [-\pi,\pi)} \Re(Z(\theta))
\]
with no constraints required.

We are thus interested in analysing the behaviour of the maximiser of $\Re(Z(\theta))$.  Recalling the definition of $R_i$ and $\Phi_i$ from~\eqref{eq:RiandPhii},
\begin{align*}
y_i &= a_0 s_i + w_i \\
&= a_0 s_i \left( 1 + \frac{w_i}{a_0 s_i} \right) \\
&= a_0 s_i R_i e^{j \Phi_i} \\
&= \rho_0 R_i e^{j ( \Phi_i + \theta_0 + \angle{s_i}) }.
\end{align*}
%The $\Phi_i$ can be identified as \emph{circular~random~variables}~\cite{Mardia_directional_statistics,Fisher1993,McKilliam_mean_dir_est_sq_arc_length2010}.  
Recalling the definition of $\hat{d}_i(\theta)$ and $\hat{u}_i(\theta)$ from~\eqref{eq:hatdfinxtheta},
\begin{align*}
\hat{u}_i(\theta) &= \round{\angle{y_i} - \theta} \\
&= \round{\theta_0 + \Phi_i + \angle{s_i} - \theta} \\
&\equiv \round{ \fracpart{\theta_0 - \theta}_{\pi} + \Phi_i + \angle{s_i}} \pmod{2\pi} \\
&= \round{ \lambda + \Phi_i + \angle{s_i} },
\end{align*}
where we put $\lambda = \fracpart{\theta_0 - \theta}_\pi$.  Because $\hat{d}_i^*(\theta) = e^{-j\hat{u}_i(\theta)}$, it follows that, when $i \in D$,
\begin{align}
 y_i e^{-j\theta} \hat{d}_i^*(\theta) &= \rho_0 R_i e^{j(\lambda + \Phi_i + \angle{s_i} - \round{\lambda + \Phi_i + \angle{s_i}})} \nonumber \\
&= \rho_0 R_i e^{j(\lambda + \Phi_i - \round{\lambda + \Phi_i})} \nonumber  \\
&= \rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}} \label{eq:yethetadhat}
\end{align}
since $\round{x + \angle{s_i}} = \round{x} + \angle{s_i}$ for all $x \in \reals$ as a result of $\angle{s_i}$ being a multiple of $\tfrac{2\pi}{M}$.  Otherwise, when $i \in P$,  
\[
y_i e^{-j\theta} p_i^* = \rho_0 R_i e^{j(\lambda + \Phi_i)}.
\]
Now,
\[
Z(\theta) = \rho_0 \sum_{i \in P} R_i e^{j(\lambda + \Phi_i)} + \rho_0  \sum_{i \in D} R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
and
\[
\Re(Z(\theta)) = \rho_0 \sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \rho_0 \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}.
\] 
Let 
\begin{equation}\label{eq:GLdefn}
G_L(\lambda) = \frac{1}{\rho_0 L} \Re(Z(\theta))
\end{equation}
and put $\hat{\lambda}_L = -\sfracpart{\hat{\theta} - \theta_0}_\pi = \sfracpart{\theta_0 - \hat{\theta}}_\pi$.  Since $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ it follows that $\hat{\lambda}_L$ is the maximiser of $G_L(\lambda)$.  We will show that $\hat{\lambda}_L$ converges almost surely to zero as $L \rightarrow \infty$.  The proof of part 1 of Theorem~\ref{thm:consistency} follows from this.

Recall the functions $G$, $h_1$ and $h_2$ defined in the statement of Theorem~\ref{thm:consistency}.  Observe that
\[
\expect G_L(\lambda) = \frac{\abs{P}}{L} h_1(\lambda) + \frac{\abs{D}}{L} h_2(\lambda)
\]
and since $\frac{\abs{P}}{L} \rightarrow p$ and $\frac{\abs{D}}{L} \rightarrow d$ as $L \rightarrow \infty$,
\[
\lim_{L \rightarrow\infty} \expect G_L(\lambda) = G(\lambda) = p h_1(\lambda)   +  d h_2(\lambda).
\]
As is customary, let $\Omega$ be the sample space on which the random variables $\{w_i\}$ are defined.  Let $A$ be the subset of the sample space $\Omega$ on which $G(\hat{\lambda}_L) \rightarrow G(0)$ as $L\rightarrow\infty$.  Lemma~\ref{lem:convtoexpGlamL} shows that $\prob\{A\} = 1$.  Let $A'$ be the subset of the sample space on which $\hat{\lambda}_L \rightarrow 0$ as $L\rightarrow \infty$.  Because $G(x)$ is uniquely maximised at $x=0$, it follows that $G(\hat{\lambda}_L) \rightarrow G(0)$ only if $\hat{\lambda}_L \rightarrow 0$ as $L \rightarrow\infty$. So $A \subseteq A'$ and therefore $\prob\{A'\} \geq \prob\{A\} = 1$.  Part 1 of Theorem~\ref{thm:consistency} follows.

@@

\begin{lemma}\label{lem:empiricprocforrho} Let $T_L(\lambda) = \expect G_L(\lambda)$.  We have
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) \big) =  X_L + o_P(1),
\]
where $X_L = \sqrt{L} \big( G_L(0) - T_L(0) \big)$.
\end{lemma}
\begin{IEEEproof}
Write
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) \big) = Y_L(\hat{\lambda}_L) + X_L
\]
where
\begin{equation}\label{eq:YLdef}
Y_L(\lambda) = \sqrt{L}\big( G_L(\lambda) - T_L(\lambda) \big) - X_L.
\end{equation}
is an empirical process indexed by $\lambda$, similar to $W_L$ from~\eqref{eq:WLdef}.  As with $W_L$ results from the literature on empirical processes can be used to show that  
%Lemma~\ref{lem:ZYLempiricproc} 
%in Section~\ref{sec:an-empirical-process} 
%shows that 
for any $\delta > 0$ and $\nu > 0$, there exists $\epsilon > 0$ such that
\[
\Pr\left\{ \sup_{\sabs{\lambda}<\epsilon}\sabs{Y_L(\lambda)} > \delta  \right\} < \nu.
\]
The proof now follows by an argument analogous to that in Lemma~\ref{lem:empiricprocc}.
\end{IEEEproof}
